What We Eval in the Shadows

### 2021-04-14 paper/main.tex
Domain Specific Language (usage)

\noindent
As shown by Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
function environment with the \c{global} kind as a distant second. This means
that most variable lookups and most side effects either read/update existing
local variables or introduce new ones. But for which function? From the point of
view of the function that called \eval, Table~\ref{tab:funoffset} gives an
offset on the call stack. Thus 0 is the direct caller and corresponds to local variables and to \emph{lexical scoping}; 1 is its parent, and
so on, and correspond to \emph{dynamic scoping}. The data suggests that in 81\% of cases, \eval access its caller's environment -- this means, the variables of the function where \eval textually
occurs are read and written to. It is interesting that some 1.5\% of sites
evaluate a code in an environment that is three frame or above from \eval. This
distance implies that, in general, modular reasoning is impossible, to
understand the behavior of any piece of code entails fully understanding the
behavior of all functions that the code may call as the actions of \eval may
happen at a distance. Finally note that any particular site may have several
kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
site has a single kind.

\begin{table}[h]
  \centering\small\hspace{-.5cm}
\begin{minipage}{3.7cm}
  \begin{tabular}{@{}r|r|r@{}}\hline
 Kind & \#sites & \%sites \\\hline
 Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
 Synthetic & \packageNbSyntheticEnvSites & \packageNbSyntheticEnvSitePercent \\
 Global &  \packageNbStrictGlobalEnvSites & \packageNbStrictGlobalEnvSitePercent \\
 Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
\end{tabular}
\caption{Kinds per site} \label{tab:highlevelenvironments}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
\begin{tabular}{@{}r|r|r@{}}\hline
 Offset & \#sites & \%sites \\\hline
  \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
  \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
  \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
  \ge 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
 \end{tabular}
\caption{Function offset}\label{tab:funoffset}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}
\begin{tabular}{@{}r|r|r@{}} \hline
Parent & \#sites & \%sites \\\hline
Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
Package & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
\end{tabular} \label{tab:newenvs}
\caption{Wrapper envs.}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
 \begin{tabular}{@{}c|c|c@{}} \hline
 \#kinds & \#sites &  \%sites \\ \hline
 \packageNbCategoryEnvira & \packageNbCategoryEnvirSitesaRnd &  \packageNbCategoryEnvirPercenta\\
 \packageNbCategoryEnvirb &  \packageNbCategoryEnvirSitesbRnd & \packageNbCategoryEnvirPercentb \\
 \packageNbCategoryEnvirc & \packageNbCategoryEnvirSitescRnd &  \packageNbCategoryEnvirPercentc\\
 \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\\hline
\end{tabular}\caption{Multiplicities}\label{tab:polyenvir}
\end{minipage}\hspace{-1cm}
\end{table}

@@

\paragraph{Discussion}
The data presented in this section is what one could expect. ({\it Function})
The majority of \eval call sites access the environment of some function on the
call stack. Usually, the current function, but sometimes frames arbitrarily far
up the call stack. One thing the data does not say is how that environment was
obtained. The expected case is that the environment was the one obtained from a
promise; as promises combine code and the environment in which that code
originates. Likely less frequent are cases where \eval is provided the results
of programmatically selecting some call stack. ({\it Synthetic}) The relative
high frequency of synthetic environments corresponds to cases when one wants to
evaluate an expression in either a restricted environment, use a data
structure as an environment. ({\it Global}) The fact that global environment
show up so frequently is likely an artifact of the way the code is run. ({\it
  Package}) There are only 92 sites that use a package environment in \eval,
this is probably best as it is bad form to add bindings or modify functions
of a loaded package. Many more sites rather wrap a package environment in a new 
environment to add variables global the package.

@@

\mypara{Symbolic differentiation.} % D and deriv

>>>>

\noindent
As shown by Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
function environment with the \c{global} kind as a distant second. This means
that most variable lookups and most side effects either read/update existing
local variables or introduce new ones. But for which function? From the point of
view of the function that called \eval, Table~\ref{tab:funoffset} gives an
offset on the call stack. Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on. The data suggests that in 81\% of cases, \eval access its caller's environment -- this means, the variables of the function where \eval textually
occurs are read and written to. It is interesting that some 1.5\% of sites
evaluate a code in an environment that is three frame or above from \eval. This
distance implies that, in general, modular reasoning is impossible, to
understand the behavior of any piece of code entails fully understanding the
behavior of all functions that the code may call as the actions of \eval may
happen at a distance. Finally note that any particular site may have several
kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
site has a single kind.

\begin{table}[h]
  \centering\small\hspace{-.5cm}
\begin{minipage}{3.7cm}
  \begin{tabular}{@{}r|r|r@{}}\hline
 Kind & \#sites & \%sites \\\hline
 Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
 Synthetic & \packageNbSyntheticEnvSites & \packageNbSyntheticEnvSitePercent \\
 Global &  \packageNbStrictGlobalEnvSites & \packageNbStrictGlobalEnvSitePercent \\
 Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
\end{tabular}
\caption{Kinds per site} \label{tab:highlevelenvironments}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
\begin{tabular}{@{}r|r|r@{}}\hline
 Offset & \#sites & \%sites \\\hline
  \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
  \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
  \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
  \ge 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
 \end{tabular}
\caption{Function offset}\label{tab:funoffset}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}
\begin{tabular}{@{}r|r|r@{}} \hline
Parent & \#sites & \%sites \\\hline
Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
Package & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
\end{tabular} 
\caption{Wrapper envs.} \label{tab:newenvs}
\end{minipage}\hspace{-.2cm}
\begin{minipage}{3.7cm}\centering
 \begin{tabular}{@{}c|c|c@{}} \hline
 \#kinds & \#sites &  \%sites \\ \hline
 \packageNbCategoryEnvira & \packageNbCategoryEnvirSitesaRnd &  \packageNbCategoryEnvirPercenta\\
 \packageNbCategoryEnvirb &  \packageNbCategoryEnvirSitesbRnd & \packageNbCategoryEnvirPercentb \\
 \packageNbCategoryEnvirc & \packageNbCategoryEnvirSitescRnd &  \packageNbCategoryEnvirPercentc\\
 \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\\hline
\end{tabular}\caption{Multiplicities}\label{tab:polyenvir}
\end{minipage}\hspace{-1cm}
\end{table}

@@

\paragraph{Discussion}
The data presented in this section is what one could expect. ({\it Function})
The majority of \eval call sites access the environment of some function on the
call stack. Usually, the current function, but sometimes frames arbitrarily far
up the call stack. One thing the data does not say is how that environment was
obtained. The expected case is that the environment was the one obtained from a
promise; as promises combine code and the environment in which that code
originates. Likely less frequent are cases where \eval is provided the results
of programmatically selecting some call stack. ({\it Synthetic}) The relative
high frequency of synthetic environments corresponds to cases when one wants to
evaluate an expression in either a restricted environment, use a data
structure as an environment. ({\it Global}) The fact that global environment
show up so frequently is likely an artifact of the way the code is run. ({\it
  Package}) There are only 92 sites that use a package environment in \eval,
this is probably best as it is bad form to add bindings or modify functions
of a loaded package. Many more sites rather wrap a package environment in a new 
environment to create singletons in a package.

@@

\mypara{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value. For instance,   base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}. They support arithmetic operations and functions on real numbers such as \c{sin}.

In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created, which contains expressions derived twice (with \c{DD}) and then typically in a list or a data frame.
\begin{lstlisting}
 expr<-expression(beta\_1 + time*beta\_2+u)
 ...
 mu = data.frame(beta\_1=0.5, beta\_2=1, time=3, u=2.3)
 ...
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}
% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd
